---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  tidy.opts = list(width.cutoff = 60), tidy = TRUE,
  dpi = 150, fig.asp = 0.5, fig.width = 7, fig.retina = 1,
  out.width = "95%",
  warning = FALSE, message = FALSE
)
```

# mixgb

<!-- badges: start -->
[![](https://img.shields.io/badge/Made%20With-R-9cf)](https://github.com/agnesdeng/mixgb)
[![](https://img.shields.io/badge/CRAN-0.1.0-9cf)](https://github.com/agnesdeng/mixgb)
[![](https://cranlogs.r-pkg.org/badges/mixgb)](https://cran.r-project.org/package=mixgb)
[![](https://img.shields.io/badge/github%20version-0.1.0-brightgreen)](https://github.com/agnesdeng/mixgb)
<!-- badges: end -->

`mixgb` is a scalable multiple imputation framework based on XGBoost, bootstrapping and predictive mean matching. The proposed framework is implemented in an R package `mixgb`. We have shown that our framework obtains less biased estimates and reflects appropriate imputation variability, while achieving high computational efficiency. For more details, please check our paper https://arxiv.org/abs/2106.01574. (Our package has been revised and updated since the preprint was posted. Simulation code in the original supplementary files may not run as expected. Revised paper and adapted code will be updated soon.)

## New updates

**May 2022**
  
  * Visual diagnostic functions of multiply imputed data
  * Use S3 instead of R6 
  * Plot functions can show masked missing data (if provided)

**April 2022**
  
  * User can now set different number of iterations `maxit`.
* Both single and multiple imputation with XGBoost can do predictive mean matching
* Bootstrap data to make `m` imputations is optional. User can set `bootstrap = FALSE` to disable bootstrap. Users can also set sampling related hyperparameters of XGBoost (subsample, colsample_bytree, colsample_bylevel, colsample_bynode) to be less than 1 to achieve similar effect.
* Add predicted mean matching type 0. Now the options for `pmm.type` are `NULL`,`0`,`1`,`2` or `"auto"` (type 2 for numeric/integer variables, no PMM for categorical variables).
* Added more validation checks
* Compatible with `data.table`
* Cross-validation to pre-tune `nrounds`

## Under development
* make m imputation parallel 
* pre-tune other hyperparameters 

## Notice
* Currently users can set XGBoost parameter `nthread` for multithreading with OpenMP support. However, MacOS has disabled OpenMP support.


## 1. Installation

You can install the development version of mixgb from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("agnesdeng/mixgb")
```

```{r}
# load mixgb
library(mixgb)
```

### 1.1 Data cleaning before imputation
It is highly recommended to clean and check your data before imputation. Here are some common issues:
  
  - Data should be a data frame.
- ID should be removed 
- Missing values should be coded as `NA` not `NaN`
- `Inf` or `-Inf` are not allowed
- Empty cells should be coded as `NA` or sensible values
- Variables of "character" type should be converted to "factor" instead
- Variables of "factor" type should have at least two levels

The function `data_clean()` can do a preliminary check and fix some obvious problems. However, it would not fix all issues related to data quality. 

```{r, eval = FALSE}
cleanWithNA.df <- data_clean(rawdata)
```

## 2. Impute missing values with `mixgb`

We first load the `mixgb` package and the `nhanes3_newborn` dataset, which contains 16 variables of various types (integer/numeric/factor/ordinal factor). There are 9 variables with missing values.

```{r}
str(nhanes3_newborn)
colSums(is.na(nhanes3_newborn))
```

To impute this dataset, we can use the default settings. The default number of imputed datasets `m = 5`. Note that we do not need to convert our data into dgCMatrix or one-hot coding format. Our package will convert it automatically. Variables should be of the following types: numeric, integer, factor or ordinal factor. 

```{r}
# use mixgb with default settings
imputed.data <- mixgb(data = nhanes3_newborn, m = 5)
```

### 2.1 Customise imputation settings

We can also customise imputation settings:

-   The number of imputed datasets
    `m`
    
-   The number of imputation iterations
    `maxit`
    
-   Whether to convert ordinal factors to integer (imputation process will be faster)
    `ordinalAsInteger`

-   Whether to use bootstrapping 
    `bootstrap`

-   Predictive mean matching settings
    `pmm.type`, `pmm.k` and `pmm.link`.

-   Initial imputation methods for different types of variables
    `initial.num`, `initial.int` and `initial.fac`.

-   Whether to save models for imputing newdata
    `save.models` and `save.vars`.
    
-   XGBoost hyperparameters and verbose settings.
    `xgb.params`, `nrounds`, `early_stopping_rounds`, `print_every_n` and `verbose`.


```{r}
# Use mixgb with chosen settings
params <- list(
  max_depth = 6,
  gamma = 0.1,
  eta = 0.3,
  min_child_weight = 1,
  subsample = 1,
  colsample_bytree = 1,
  colsample_bylevel = 1,
  colsample_bynode = 1,
  nthread = 4,
  tree_method = "auto",
  gpu_id = 0,
  predictor = "auto"
)

imputed.data <- mixgb(
  data = nhanes3_newborn, m = 5, maxit = 1,
  ordinalAsInteger = TRUE, bootstrap = TRUE,
  pmm.type = "auto", pmm.k = 5, pmm.link = "prob",
  initial.num = "normal", initial.int = "mode", initial.fac = "mode",
  save.models = FALSE, save.vars = NULL,
  xgb.params = params, nrounds = 50, early_stopping_rounds = 10, print_every_n = 10L, verbose = 0
)
```


### 2.2 Tune hyperparameters

Imputation performance can be affected by the hyperparameter settings. It may seem daunting to tune a large set of hyperparameters, but often we can narrow down the search as many hyperparameters are correlated. In our package, we have a function `mixgb_cv()` to tune `nrounds`. There is no default `nrounds` value in `XGBoost,` so we need to specify it. The default `nrounds` in `mixgb` is 50. However, we recommend using `mixgb_cv()` to find the optimal `nrounds` first.

```{r}
cv.results <- mixgb_cv(data = nhanes3_newborn, verbose = FALSE)
cv.results$response
cv.results$best.nrounds
```

By default, `mixgb_cv()` will randomly choose an incomplete variable as the response and build an XGBoost model with other variables using the complete cases of the dataset. Therefore, each run of `mixgb_cv()` is likely to return different results. Users can also specify the response and covariates in the argument `response` and `select_features`, respectively.

```{r}
cv.results <- mixgb_cv(data = nhanes3_newborn, nfold = 10, nrounds = 100, early_stopping_rounds = 1, 
                       response = "BMPHEAD", select_features = c("HSAGEIR", "HSSEX", "DMARETHN", "BMPRECUM",                           "BMPSB1", "BMPSB2", "BMPTR1", "BMPTR2", "BMPWT"), verbose = FALSE)

cv.results$best.nrounds
```

Since using `mixgb_cv()` with this dataset mostly returns a number less than `20`, I'll set `nrounds = 20` in `mixgb()` to obtain `m` imputed datasets. 

```{r, eval = FALSE}
imputed.data <- mixgb(data = nhanes3_newborn, m = 5, nrounds = 20)
```


## 3. Visualize multiply imputed values
It is important to assess the plausibility of imputations before doing analysis. The `mixgb` package provides several visual diagnostic functions using `ggplot2` to compare multiply imputed values versus observed data.

We will demonstrate these functions using the `nhanes3_newborn` dataset. In the original data, almost all missing values occurred in numeric variables. Only seven observations are missing in the binary factor variable `HFF1` . In order to visualize some imputed values for other types of variables, we create some extra missing values in `HSHSIZER` (integer), `HSAGEIR` (integer), `HSSEX` (binary factor), `DMARETHN` (multiclass factor) and `HYD1` (Ordinal factor) under MCAR.

```{r}
withNA.df <- createNA(data = nhanes3_newborn, var.names = c("HSHSIZER", "HSAGEIR", "HSSEX", "DMARETHN", "HYD1"), p = 0.1)
colSums(is.na(withNA.df))
```
We then impute this dataset using `mixgb()` with default settings. A list of five imputed datasets are assigned to `imputed.data`. The dimension of each imputed dataset will be the same as the original data.
```{r}
imputed.data <- mixgb(data = withNA.df, m = 5)
```

The `mixgb` package provides the following visual diagnostics functions:

(i) Single variable: `plot_hist()`, `plot_box()`, `plot_bar()` ;

(ii) Two variables: `plot_2num()`, `plot_2fac()`, `plot_1num1fac()` ;

(iii) Three variables: `plot_2num1fac()`, `plot_1num2fac()`.

Each function will return `m+1` panels to compare the observed data with `m` sets of actual imputed values.

Here are some examples. For more details, please check the vignette [Visual diagnostics for multiply imputed values][Visual-diagnostics].
```{r}
plot_hist(imputation.list = imputed.data, var.name = "BMPHEAD", original.data = withNA.df)
plot_2num(imputation.list = imputed.data, var.x = "BMPHEAD", var.y = "BMPRECUM", original.data = withNA.df)
plot_2num(imputation.list = imputed.data, var.x = "HSAGEIR", var.y = "BMPHEAD", original.data = withNA.df)
plot_1num1fac(imputation.list = imputed.data, var.num = "BMPHEAD", var.fac = "HSSEX", original.data = withNA.df)
```

## 4. Impute new unseen data using a saved imputer object

First we can split the `nhanes3_newborn` dataset into training data and test data.

```{r}
library(mixgb)
data("nhanes3_newborn")
set.seed(2022)
n <- nrow(nhanes3_newborn)
idx <- sample(1:n, size = round(0.7 * n), replace = FALSE)
train.data <- nhanes3_newborn[idx, ]
test.data <- nhanes3_newborn[-idx, ]
```

We can use the training data to obtain `m` imputed datasets and save their imputation models. To achieve this, users need to set `save.models = TRUE`. By default `save.vars = NULL`, imputation models for variables with missing data in the training data will be saved. However, the unseen data may also have missing values in other variables. Users can be comprehensive by saving models for all variables by setting `save.vars = colnames(train.data)`. Note that this would take much longer as we need to train and save a model for each variable. If users are confident that only certain variables will have missing values in the new data, we recommend specifying the names or indices of these variables in `save.vars` instead of saving models for all variables.

```{r}
# obtain m imputed datasets for train.data and save imputation models
mixgb.obj <- mixgb(data = train.data, m = 5, save.models = TRUE, save.vars = NULL)
```

When `save.models = TRUE`, `mixgb()` will return an object containing the following:

-   `imputed.data`: a list of `m` imputed dataset for training data

-   `XGB.models`: a list of `m` sets of XGBoost models for variables specified in `save.vars`.

-   `params`: a list of parameters that are required for imputing new data using `impute_new()` later on.

We can extract `m` imputed datasets from the saved imputer object by `$imputed.data`.

```{r}
train.imputed <- mixgb.obj$imputed.data
# the 5th imputed dataset
head(train.imputed[[5]])
```

To impute new data with this saved imputer object, we use the `impute_new()` function. User can also specify whether to use new data for initial imputation. By default, `initial.newdata = FALSE`, we will use the information of training data to initially impute the new data. New data will be imputed with the saved models. This process will be considerably faster as we don't need to build the imputation models again.

```{r}
test.imputed <- impute_new(object = mixgb.obj, newdata = test.data)
```

If PMM is used when we call `mixgb()`, predicted values of missing entries in the new dataset are matched with donors from training data. Users can also set the number of donors for PMM when imputing new data. By default, `pmm.k = NULL` , which means the same setting as the training object will be used.

Similarly, users can set the number of imputed datasets `m`. Note that this value has to be smaller than or equal to the `m` in `mixgb()`. If it is not specified, it will use the same `m` value as the saved object.

```{r}
test.imputed <- impute_new(object = mixgb.obj, newdata = test.data, initial.newdata = FALSE, pmm.k = 3, m = 4)
```


## 5. Install `mixgb` with GPU support
Multiple imputation can be run with GPU support for machines with NVIDIA GPUs. Note that users have to install the R package `xgboost` with GPU support first. 

The `xgboost` R package pre-built binary on Linux x86_64 with GPU support can be downloaded from the release page https://github.com/dmlc/xgboost/releases/tag/v1.4.0

The package can then be installed by running the following commands:

``` 
# Install dependencies
$ R -q -e "install.packages(c('data.table', 'jsonlite'))"

# Install XGBoost
$ R CMD INSTALL ./xgboost_r_gpu_linux.tar.gz
```

Then users can install package `mixgb` in R. 

```{r, eval = FALSE}
devtools::install_github("agnesdeng/mixgb")
library(mixgb)
```
Users just need to specify ` tree_method = "gpu_list"` in the params list which will then be passed to `xgb.params` in `mixgb()`. Other GPU-realted arguments include  `gpu_id`  and  `predictor`. By default, `gpu_id = 0` and `predictor = "auto"`.

```{r, eval = FALSE}
params <- list(
  max_depth = 6,
  gamma = 0.1,
  eta = 0.3,
  min_child_weight = 1,
  subsample = 1,
  colsample_bytree = 1,
  colsample_bylevel = 1,
  colsample_bynode = 1,
  nthread = 4,
  tree_method = "gpu_list",
  gpu_id = 0,
  predictor = "auto"
)


mixgb.data <- mixgb(data = withNA.df, m = 5, xgb.params = params)
```

[Visual-diagnostics]:https://agnesdeng.github.io/mixgb/articles/web/Visual-diagnostics.html
