---
title: "mixgb: Multiple Imputation Through XGBoost"
author: "Agnes Yongshi Deng"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mixgb: Multiple Imputation Through XGBoost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Imputing missing values with `mixgb`

We first load the `mixgb` package and the `nhanes3_newborn` dataset, which contains 16 variables of various types (integer/numeric/factor/ordinal factor). There are 9 variables with missing values.

```{r setup}
library(mixgb)
data(nhanes3_newborn)
str(nhanes3_newborn)
colSums(is.na(nhanes3_newborn))
```

To impute this dataset, first we need to set up an Mixgb imputer with our choice of settings or leave it as default. Note that we do not need to convert our data into dgCMatrix or one-hot coding format. Our package will convert it automatically. Variables should be of the following types: numeric, integer, factor or ordinal factor. Except `data`, Other arguments in `Mixgb$new( )` are either hyperparameters or verbose settings for XGBoost training.

```{r}
#default
MIXGB <- Mixgb$new(data = nhanes3_newborn)

#with chosen settings
params = list(max_depth = 6, 
              gamma = 0.1, 
              eta = 0.3, 
              min_child_weight = 1, 
              subsample = 1, 
              colsample_bytree = 1, 
              colsample_bylevel = 1, 
              colsample_bynode=1, 
              nthread = 4, 
              tree_method = "auto", 
              gpu_id = 0, 
              predictor = "auto")

MIXGB <- Mixgb$new(data = nhanes3_newborn, xgb.params = params, nrounds = 100, early_stopping_rounds = 1, print_every_n = 1, verbose = 0)
```

### Tuning hyperparameters

Imputation performance can be affected by the Hyperparameters settings. It may seems daunting to tune a large set of hyperparameters, but most of the time, we can narrow down the search as many hyperparameters are correlated. There is no default `nrounds` value in `XGBoost` and users need to specify it. In our package `mixgb`, we set the default `nrounds` to be 50. However, we recommend users to use `mixgb_cv()` to find the optimal `nrounds` first.

```{r}
optimal.nrounds <- mixgb_cv(data = nhanes3_newborn, verbose = FALSE)
optimal.nrounds$evaluation.log
optimal.nrounds$best.nrounds
```

By default, `mixgb_cv()` will randomly choose an incomplete variable as the response and build an XGBoost model with other variables using the complete cases of the dataset. Therefore, each run of `mixgb_cv()` are likely to return different results. Users can also specify the response and covariates in the argument `response` and `select_covariates`, respectively.

```{r}
optimal.nrounds <- mixgb_cv(data = nhanes3_newborn, nfold = 10, nrounds = 100, early_stopping_rounds = 1, response = "BMPHEAD", select_covariates = c("HSHSIZER","HSAGEIR", "HSSEX", "DMARETHN", "BMPRECUM", "BMPSB1", "BMPSB2", "BMPTR1", "BMPTR2"  , "BMPWT", "DMPPIR", "HFF1", "HYD1"),verbose = FALSE)
optimal.nrounds$best.nrounds
```

Since the results returned by `mixgb_cv()` are mostly less than 20, I'll set `nrounds = 20` in the imputer.

```{r}
MIXGB <- Mixgb$new(data = nhanes3_newborn, nrounds = 20)
```

We now use this imputer to obtain m imputed datasets, which will be stored in a list of `m` data tables. If `m = 1`, single imputation will be undertaken. Users can also specify the number of imputation iterations `maxit`.

```{r}
imputed.data <- MIXGB$impute(m = 5, maxit = 1)
```

We can also customise other imputation settings:

-   Whether to use bootstrapping

-   The types of predictive mean matching

-   The number of donors for predictive mean matching

-   Whether to convert ordinal factors to integer ( imputation process will be faster)

-   Initial imputation methods

```{r}
imputed.data <- MIXGB$impute(m = 5, maxit = 1,  bootstrap = TRUE,
                             pmm.type = "auto", pmm.k = 5, 
                             ordinalAsInteger = TRUE,
                            initial.num = "normal", initial.int = "sample", initial.fac = "mode")

```

### Convert ordinal factors to integers

If our dataset has a decent amount of ordinal factors, imputation process will speed up if we convert them to integers. In `nhanes3_newborn` data, there is only one ordinal factor, but we can still see that the imputation time is smaller with `ordinalAsInteger = TRUE`. If `m` and `maxit` are larger, the difference will be even more obvious. Note that when `ordinalAsInteger = TRUE` , the class of these ordinal factors in the imputed datasets will also be integer. We recommend to use the default PMM setting `pmm.type = "auto"`, which will impute numeric and integer variables with PMM (type 2) and impute categorical variables without PMM. If we don't apply PMM to integer variables, the missing values may be imputed as numeric. Sometimes this can be problematic, especially when the integers are actually converted from ordinal factors. Therefore, `pmm.type = NULL` is disabled when `ordinalAsInteger = TRUE` .

```{r}
system.time(imputed.data <- MIXGB$impute(m = 5, maxit = 1,ordinalAsInteger = FALSE))

system.time(imputed.data <- MIXGB$impute(m = 5, maxit = 1,ordinalAsInteger = TRUE))

```

After imputing, we can have a look at the imputed values for an variable of interest. For example, here are 5 imputed sets of the missing values of the variable `BMPHEAD`.

```{r}
imputed.BMPHEAD<-imputed_var(imputation.list = imputed.data, var.name = "BMPHEAD", original.data = nhanes3_newborn)
imputed.BMPHEAD
```

## Impute new unseen data using a trained imputer
First we can split the `nhanes3_newborn` dataset into training data and test data.
```{r}
set.seed(2022)
n <- nrow(nhanes3_newborn)
idx <- sample(1:n, size = round(0.7 * n), replace = FALSE)
train.data <- nhanes3_newborn[idx, ]
test.data <- nhanes3_newborn[-idx, ]
```

We can use the training data to obtain `m` imputed datasets and save their imputation models. To achieve this, users need to set `save.models = TRUE`. By default `save.vars = NULL`, imputation models for variables with missing data in the training data will be saved. However, the unseen data may also have missing values in other variables, it would be comprehensive to save models for all variables by setting `save.vars = colnames(train.data)`. Note that this would take much longer time as we need to train and save a model for each variable. If users are confident that only certain variables will have missing values in the new data, we recommend to specify the names or indices of these variables in `save.vars` instead of saving models for all variables.

```{r}
MIXGB <- Mixgb$new(data = train.data)
mixgb.obj <- MIXGB$impute(m = 5, maxit = 1, save.models = TRUE, save.vars = NULL)
```

We can extract `m` imputed datasets from the saved imputer object by `$imputed.data`.

```{r}
imputed.traindata <- mixgb.obj$imputed.data
#the 5th imputed dataset
head(imputed.traindata[[5]])
```

To impute new data with this saved imputer object, we use the `impute_new()` function. User can also specify whether to use new data for initial imputation. By default, `initial.newdata = FALSE`, we will use the information of training data to initially impute the new data. New data will be imputed with the saved models. This process will be considerably fast as we don't need to build the imputation models again.

```{r}
imputed.testdata <- impute_new(object = mixgb.obj, newdata = test.data) 

```

If PMM is used in the trained imputer, predicted values of missing entries in the new dataset are matched with training data. Users can also set the number of donors for PMM when imputing new data. By default, `pmm.k = NULL` , which means the same setting as the training object will be used.

Similarly, users can set the number of imputed datasets `m`. Note that this value has to be smaller than or equal to the one set in the trained imputer. If it is not specified, it will use the same `m` value as the trained imputer.

```{r}
imputed.testdata <- impute_new(object = mixgb.obj, newdata = test.data, initial.newdata = FALSE, pmm.k = 3, m = 4)
```
