---
title: "mixgb: Multiple Imputation Through XGBoost"
author: "Yongshi Deng"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{mixgb: Multiple Imputation Through XGBoost}
  %\VignetteEngine{knitr::rmarkdown}
---
## Introduction

Mixgb offers a scalable solution for imputing large datasets using
XGBoost, bootstrapping and predictive mean matching. Mixgb is built
under Fully Conditional Specification (FCS), where XGBoost imputation
models are built for each incomplete variable. Mixgb can automatically
handle different types of variables and users do not need to encode
categorical variables themselves. Users can also choose different
settings regarding bootstrapping and the types predictive mean matching
to enhance imputation performance.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Impute missing values with `mixgb`

We first load the `mixgb` package and the `nhanes3_newborn` dataset,
which contains 16 variables of various types
(integer/numeric/factor/ordinal factor). There are 9 variables with
missing values.

```{r}
library(mixgb)
str(nhanes3_newborn)
colSums(is.na(nhanes3_newborn))
```

To impute this dataset, we can use the default settings. The default
number of imputed datasets `m = 5`. Note that we do not need to convert
our data into dgCMatrix or one-hot coding format. Our package will
convert it automatically. Variables should be of the following types:
numeric, integer, factor or ordinal factor.

```{r, eval = FALSE}
# use mixgb with default settings
imputed.data <- mixgb(data = nhanes3_newborn, m = 5)
```

### Customise imputation settings

We can also customise imputation settings:

-   The number of imputed datasets `m`

-   The number of imputation iterations `maxit`

-   Whether to convert ordinal factors to integer (imputation process
    will be faster) `ordinalAsInteger`

-   Whether to use bootstrapping `bootstrap`

-   Predictive mean matching settings `pmm.type`, `pmm.k` and
    `pmm.link`.

-   Initial imputation methods for different types of variables
    `initial.num`, `initial.int` and `initial.fac`.

-   Whether to save models for imputing newdata `save.models` and
    `save.vars`.

-   XGBoost hyperparameters and verbose settings. `xgb.params`,
    `nrounds`, `early_stopping_rounds`, `print_every_n` and `verbose`.

```{r, eval = FALSE}
# Use mixgb with chosen settings
params <- list(
  max_depth = 6,
  gamma = 0.1,
  eta = 0.3,
  min_child_weight = 1,
  subsample = 1,
  colsample_bytree = 1,
  colsample_bylevel = 1,
  colsample_bynode = 1,
  nthread = 4,
  tree_method = "auto",
  gpu_id = 0,
  predictor = "auto"
)

imputed.data <- mixgb(
  data = nhanes3_newborn, m = 5, maxit = 1,
  ordinalAsInteger = TRUE, bootstrap = TRUE,
  pmm.type = "auto", pmm.k = 5, pmm.link = "prob",
  initial.num = "normal", initial.int = "mode", initial.fac = "mode",
  save.models = FALSE, save.vars = NULL,
  xgb.params = params, nrounds = 50, early_stopping_rounds = 10, print_every_n = 10L, verbose = 0
)
```

### Tune hyperparameters

Imputation performance can be affected by the hyperparameter settings.
It may seem daunting to tune a large set of hyperparameters, but often
we can narrow down the search as many hyperparameters are correlated. In
our package, we have a function `mixgb_cv()` to tune `nrounds`. There is
no default `nrounds` value in `XGBoost,` so we need to specify it. The
default `nrounds` in `mixgb` is 50. However, we recommend using
`mixgb_cv()` to find the optimal `nrounds` first.

```{r}
cv.results <- mixgb_cv(data = nhanes3_newborn, verbose = FALSE)
cv.results$evaluation.log
cv.results$response
cv.results$best.nrounds
```

By default, `mixgb_cv()` will randomly choose an incomplete variable as
the response and build an XGBoost model with other variables using the
complete cases of the dataset. Therefore, each run of `mixgb_cv()` is
likely to return different results. Users can also specify the response
and covariates in the argument `response` and `select_features`,
respectively.

```{r}
cv.results <- mixgb_cv(data = nhanes3_newborn, nfold = 10, nrounds = 100, early_stopping_rounds = 1, 
                       response = "BMPHEAD", select_features = c("HSAGEIR", "HSSEX", "DMARETHN", "BMPRECUM",                           "BMPSB1", "BMPSB2", "BMPTR1", "BMPTR2", "BMPWT"), verbose = FALSE)

cv.results$best.nrounds
```

Since using `mixgb_cv()` with this dataset mostly returns a number less
than `20`, I'll set `nrounds = 20` in `mixgb()` to obtain `m` imputed
datasets.

```{r, eval = FALSE}
imputed.data <- mixgb(data = nhanes3_newborn, m = 5, nrounds = 20)
```

#### Convert ordinal factors to integers

If our dataset has ordinal factors, then the imputation process will be
sped up by converting them to integers. In `nhanes3_newborn` data, there
is only one ordinal factor, but we can still see that the imputation
time is smaller with `ordinalAsInteger = TRUE`. If `m` and `maxit` are
larger, the difference will be even more obvious. Note that when
`ordinalAsInteger = TRUE` , the class of these ordinal factors in the
imputed datasets will also be integer. We recommend to use the default
PMM setting `pmm.type = "auto"`, which will impute numeric and integer
variables with PMM (type 2) and impute categorical variables without
PMM. If we don't apply PMM to integer variables, the missing values may
be imputed as numeric. Sometimes this can be problematic, especially
when the integers are actually converted from ordinal factors.
Therefore, `pmm.type = NULL` is disabled when `ordinalAsInteger = TRUE`
.

```{r}
system.time(imputed.data <- mixgb(data = nhanes3_newborn, m = 5, maxit = 1, ordinalAsInteger = FALSE))

system.time(imputed.data <- mixgb(data = nhanes3_newborn, m = 5, maxit = 1, ordinalAsInteger = TRUE))
```

## Inspect multiply imputed values

The `mixgb` package provides the following visual diagnostics functions:

(i) Single variable: `plot_hist()`, `plot_box()`, `plot_bar()` ;

(ii) Two variables: `plot_2num()`, `plot_2fac()`, `plot_1num1fac()` ;

(iii) Three variables: `plot_2num1fac()`, `plot_1num2fac()`.

Each function will return `m+1` panels to compare the observed data with
`m` sets of actual imputed values.

For more details, please check the vignette on github[Visual diagnostics
for multiply imputed
values](https://agnesdeng.github.io/mixgb/articles/web/Visual-diagnostics.html).
