---
title: "Visual diagnostics for multiply imputed values"
author: "Yongshi Deng"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Visual diagnostics for multiply imputed values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy.opts = list(width.cutoff = 60), tidy = TRUE,
  dpi = 200, fig.asp = 0.5, fig.width = 7, fig.retina = 1,
  out.width = "95%",
  warning = FALSE, message = FALSE
)
```

# 1. Introduction

It is crucial to assess the plausibility of imputations before doing an analysis. The `mixgb` package provides several visual diagnostic functions using `ggplot2` to compare multiply imputed values versus observed data.

## 1.1 Inspecting imputed values

We will demonstrate these functions using the `nhanes3_newborn` dataset. In the original data, almost all missing values occurred in numeric variables. Only seven observations are missing in the binary factor variable `HFF1`. 

```{r}
library(mixgb)
colSums(is.na(nhanes3_newborn))
```

In order to visualize some imputed values for other types of variables, we create some extra missing values in `HSHSIZER` (integer), `HSAGEIR` (integer), `HSSEX` (binary factor), `DMARETHN` (multiclass factor) and `HYD1` (Ordinal factor) under the missing completely at random (MCAR) mechanism.

```{r}
withNA.df <- createNA(data = nhanes3_newborn, var.names = c("HSHSIZER", "HSAGEIR", "HSSEX", "DMARETHN", "HYD1"), p = 0.1)
colSums(is.na(withNA.df))
```

We then impute this dataset using `mixgb()` with default settings. After completion, a list of five imputed datasets is assigned to `imputed.data`. Each imputed dataset will have the same dimension as the original data.

```{r}
imputed.data <- mixgb(data = withNA.df, m = 5)
```

By using the function `show_var()`, we can see the multiply imputed values of the missing data for a given variable. The function returns a data.table of `m` columns, each of which represents a set of imputed values for the variable of interest. Note that the output of this function only includes the imputed values of missing entries in the specified variable.

```{r}
show_var(imputation.list = imputed.data, var.name = "BMPHEAD", original.data = withNA.df)
show_var(imputation.list = imputed.data, var.name = "HFF1", original.data = withNA.df)
```

# 2 Visual diagnostics plots

The `mixgb` package provides the following visual diagnostics functions:

(i) Single variable: `plot_hist()`, `plot_box()`, `plot_bar()` ;

(ii) Two variables: `plot_2num()`, `plot_2fac()`, `plot_1num1fac()` ;

(iii) Three variables: `plot_2num1fac()`, `plot_1num2fac()`.

Each function returns `m+1` panels that enable a comparison between the observed data and `m` sets of imputed values for the missing data.

## 2.1 Single variable

Only imputed values of missing entries in the specified variable will be plotted in panels `m1` to `m5`. An error message will be displayed if the variable is fully observed and there are no missing entries to impute.

### 2.1.1 Numeric

We can plot an imputed numeric variable by histogram or boxplot.

-   `plot_hist()`: plot histograms with density curves.

    Histograms are a useful tool for displaying the distribution of numeric data. Users can examine the shapes of the histograms to identify any unusual patterns in the imputed values. If the data is missing completely at random (MCAR), we would expect the distribution of imputed values to be the same as that of the observed values. However, if the data is missing at random (MAR), the distributions of observed and imputed values can be quite different. Nevertheless, it is still worth plotting the imputed data, as any unusual values may indicate that the imputation procedure is unsatisfactory.

    ```{r}
plot_hist(
  imputation.list = imputed.data, var.name = "BMPHEAD",
  original.data = withNA.df
)
    ```

-   `plot_box()`: plot box plots with overlaying data points.

    Users can use `plot_box()` to compare the median, lower and upper quantiles of imputed values with those of the observed values. The plot also visually displays the disparity in counts between observed and missing values in the variable of interest.

    ```{r}
plot_box(
  imputation.list = imputed.data, var.name = "BMPHEAD",
  original.data = withNA.df
)
    ```

### 2.1.2 Factor

-   `plot_bar()`: plot bar plots

    The proportion of each level in a factor will be shown by `plot_bar()`.

    ```{r}
plot_bar(
  imputation.list = imputed.data, var.name = "HSSEX",
  original.data = withNA.df
)
plot_bar(
  imputation.list = imputed.data, var.name = "DMARETHN",
  original.data = withNA.df
)
    ```

### 2.1.3 Integer

Users have the flexibility to choose how to treat integer variables when generating a plot, as they can be viewed either as numeric or factor variables. To plot an imputed integer variable, users can use one of the following functions based on their preferred representation:

-   `plot_hist()`: plot histograms with density curves

-   `plot_box()`: plot box plot with overlaying data points

-   `plot_bar()`: plot bar plot (treat an integer variable as a factor)

    ```{r}
plot_hist(
  imputation.list = imputed.data, var.name = "HSHSIZER",
  original.data = withNA.df
)
plot_box(
  imputation.list = imputed.data, var.name = "HSHSIZER",
  original.data = withNA.df
)
plot_bar(
  imputation.list = imputed.data, var.name = "HSHSIZER",
  original.data = withNA.df
)
    ```

### 2.1.4 Ordinal factor

By default, the function `mixgb()` does not convert ordinal factors to integers. Therefore, we may simply plot ordinal factors as factors (see Section 2.1.2).

However, setting `ordinalAsInteger = TRUE` in `mixgb()` may speed up the imputation process, but users must decide whether to transform them back. If we choose to convert ordinal factors to integers prior to imputation, the imputed values must be plotted as if they were integers (see Section 2.1.3).


```{r}
imputed.data2 <- mixgb(data = withNA.df, m = 5, ordinalAsInteger = TRUE)

plot_bar(
  imputation.list = imputed.data2, var.name = "HYD1",
  original.data = withNA.df
)
plot_hist(
  imputation.list = imputed.data2, var.name = "HYD1",
  original.data = withNA.df
)
plot_box(
  imputation.list = imputed.data2, var.name = "HYD1",
  original.data = withNA.df
)
```

## 2.2 Two variables

All of the functions in this section for visualizing the multiply imputed values of two variables require at least one of the variables in the original dataset to be incomplete. The imputed values shown in panels `m1` to `m5` are those that were originally missing in one or both of the variables.

### 2.2.1 Two numeric variables

We can generate scatter plots of two imputed numeric variables by using `plot_2num()`. We can specify the x-axis variable in `var.x` and the y-axis variable in `var.y`.

Users can choose to plot the shapes of different types of missing values by setting `shape = TRUE`. We only recommend plotting the shapes for small datasets. By default, `shape = FALSE` is used to expedite the plotting process.

```{r}
plot_2num(
  imputation.list = imputed.data, var.x = "BMPHEAD", var.y = "BMPRECUM",
  original.data = withNA.df, shape = TRUE
)
```

**NA.condition** represents the following four types of missing values.

(i) `both.observed`: Both `var.x` and`var.y` are observed. This only appears in the first panel - `Observed` (Shape: diamond).

(ii) `both.missing`: Imputed values where both `var.x` and `var.y` are originally missing (Shape: circle);

(iii) `var.x.missing`: Imputed values where `var.x` is originally missing and `var.y` is not (Shape: X);

(iv) `var.y.missing`: Imputed values where `var.y` is originally missing and `var.x` is not (Shape: Y).

### 2.2.2 One numeric vs one factor

We can plot a numeric variable versus a factor using `plot_1num1fac()`. The output of this function is a box plot with overlaying points. Users are required to specify a numeric variable in `var.num` and a factor in `var.fac`.

**NA.condition** is similar to the definition in Section 2.2.1.

```{r}
plot_1num1fac(
  imputation.list = imputed.data, var.num = "BMPHEAD", var.fac = "HSSEX",
  original.data = withNA.df
)
```

### 2.2.3 Two factors

We can generate bar plots to show the relationship between two factors by using `plot_2fac()`. 

```{r}
plot_2fac(
  imputation.list = imputed.data, var.fac1 = "HYD1", var.fac2 = "HFF1",
  original.data = withNA.df
)
```

### 2.2.4 One numeric vs one integer

We can use `plot_2num()` to visualize the relationship between a numeric variable and an integer variable. Note that the graphs would appear differently if the variables `var.x` and `var.y` were swapped.

```{r}
plot_2num(
  imputation.list = imputed.data, var.x = "BMPHEAD", var.y = "HSAGEIR",
  original.data = withNA.df, shape = TRUE
)
plot_2num(
  imputation.list = imputed.data, var.x = "HSAGEIR", var.y = "BMPHEAD",
  original.data = withNA.df, shape = TRUE
)
```

If the integer variable `HSAGEIR` is to be treated as a factor rather than a numeric variable, the function `plot 1num1fac()` can be used.

```{r}
plot_1num1fac(
  imputation.list = imputed.data, var.num = "BMPHEAD", var.fac = "HSAGEIR",
  original.data = withNA.df, shape = TRUE
)
```

### 2.2.5 Two integers

We can plot two integer variables using either `plot_2num()` ,`plot_1num1fac()`, or `plot_2fac()`. Users should choose the plotting functions based on the nature of the variable. For example, if an integer variable `age` has values between 0 and 110, it may be more convenient to treat `age` as numeric rather than a factor. If, on the other hand, an integer variable has just a few unique values (such as 1, 2, 3), it may be preferable to be plotted as a factor.  In this dataset, there are only two variables of integer type - `HSHSIZER`(household size) and `HSAGEIR`(baby's age ranging from 2 to 11 months). Although there is no expected relationship between these two, we still create a plot for demonstration purposes.

```{r}
plot_2num(
  imputation.list = imputed.data, var.x = "HSHSIZER", var.y = "HSAGEIR",
  original.data = withNA.df, shape = TRUE
)
plot_1num1fac(
  imputation.list = imputed.data, var.num = "HSHSIZER", var.fac = "HSAGEIR",
  original.data = withNA.df, shape = TRUE
)
plot_2fac(
  imputation.list = imputed.data, var.fac1 = "HSHSIZER", var.fac2 = "HSAGEIR",
  original.data = withNA.df
)
```

## 2.3 Three variables

To plot the multiply imputed values of three variables, at least one of them has to be incomplete in the original dataset.

### 2.3.1 Two numeric variables conditional on one factor

We can generate a scatter plot of two numeric variables when conditioned on a factor by using `plot_2num1fac()`. The variable for the x-axis should be specified in `var.x`, whereas the variable for the y-axis should be specified in `var.y`. The factor on which we want to condition is `con.fac`.

```{r}
plot_2num1fac(
  imputation.list = imputed.data, var.x = "BMPHEAD", var.y = "BMPRECUM",
  con.fac = "HFF1", original.data = withNA.df
)
```

When we have three variables, there are $2^3$ different types of missing patterns. These patterns consist of all possible combinations of zero to three variables missing. However, attempting to differentiate all eight types of missingness in the same plot can be challenging, particularly when working with large datasets. Therefore, we have decided to display the following three types of missing patterns (**NA.condition**) in the plot when `shape = TRUE`.

**NA.condition** represents the following three types of missing values.

(i) `all.observed`: Observations where all three variables are observed. This only appears in the first panel - `Observed` .

(ii) `con.fac.observed`: Imputed values where `con.fac` is originally observed.

     (These points are originally missing in either `var.x` or `var.y` or both)

(iii) `con.fac.missing`: imputed values where `con.fac` is originally missing. (These points can be originally observed, or missing in either `var.x` or `var.y` or both)

```{r}
plot_2num1fac(
  imputation.list = imputed.data, var.x = "BMPHEAD", var.y = "BMPRECUM",
  con.fac = "DMARETHN", original.data = withNA.df, shape = TRUE
)
```

If we want to treat an integer variable as numeric, we can place it in either `var.x` or `var.y`. Here is an example, where `HSAGEIR` is an integer variable whose values range from 2 to 11.

```{r,fig.asp = 1}
plot_2num1fac(
  imputation.list = imputed.data, var.x = "HSAGEIR", var.y = "BMPRECUM",
  con.fac = "DMARETHN", original.data = withNA.df
)
```

### 2.3.2 One numeric variable and one factor conditional on another factor

The function `plot_1num2fac()` generates boxplots with overlaying data points for a numeric variable with a factor, conditional on another factor.

```{r,fig.asp = 0.7}
plot_1num2fac(
  imputation.list = imputed.data, var.fac = "DMARETHN", var.num = "BMPRECUM",
  con.fac = "HSSEX", original.data = withNA.df
)
```

# 3 Color options

By default, the observed panel is gray and the other `m` panels use `ggplot2`'s default color scheme.

```{r}
plot_2num(
  imputation.list = imputed.data, var.x = "BMPHEAD", var.y = "BMPRECUM",
  original.data = withNA.df, color.pal = NULL
)
```

The `color.pal` argument allows users to specify a custom color palette with a vector of hex codes, such as the colorblind-friendly palette `Set2` from the R package `RColorBrewer`. Note that if there are `m` imputed datasets, a vector of `m+1` hex codes is required because an extra color is needed for the `Observed` panel.

```{r}
library(RColorBrewer)
color.codes <- brewer.pal(n = 6, name = "Set2")
color.codes

plot_2num(
  imputation.list = imputed.data, var.x = "BMPHEAD", var.y = "BMPRECUM",
  original.data = withNA.df, color.pal = color.codes
)
```

Otherwise, we can provide a vector of R's built-in color names.

```{r}
color.names <- c("gray50", "coral2", "goldenrod3", "darkolivegreen4", "slateblue1", "plum3")

plot_2num(
  imputation.list = imputed.data, var.x = "BMPHEAD", var.y = "BMPRECUM",
  original.data = withNA.df, color.pal = color.names
)
```

Here is a very useful cheat sheet of R colors names created by Dr Ying Wei.

[http://www.stat.columbia.edu/\\\~tzheng/files/Rcolor.pdf](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf){.uri}

# 4 Plot multiply imputed data from other packages

We can also plot multiply imputed datasets obtaining from other packages, such as `mice`. Here is an example using the `nhanes2` data from `mice`.

This dataset consists of just 25 rows and 4 columns (`age`, `bmi`, `hyp` and `chl`). There are only 9, 8 and 10 missing values in the variables `bmi`, `hyp`, and `chl`, respectively. Note that imputed values are volatile when the dataset is this small.

```{r}
library(mice)
dim(nhanes2)
colSums(is.na(nhanes2))

imp <- mice(data = nhanes2, m = 5, printFlag = FALSE)
mice.data <- complete(imp, action = "all")
```

```{r}
plot_hist(imputation.list = mice.data, var.name = "bmi", original.data = nhanes2)
plot_box(imputation.list = mice.data, var.name = "chl", original.data = nhanes2)
plot_bar(imputation.list = mice.data, var.name = "hyp", original.data = nhanes2)
plot_2num(
  imputation.list = mice.data, var.x = "bmi", var.y = "chl",
  original.data = nhanes2
)
plot_1num1fac(
  imputation.list = mice.data, var.num = "chl", var.fac = "hyp",
  original.data = nhanes2
)
plot_2num1fac(
  imputation.list = mice.data, var.x = "chl", var.y = "bmi",
  con.fac = "age", original.data = nhanes2
)
plot_2num1fac(
  imputation.list = mice.data, var.x = "chl", var.y = "bmi",
  con.fac = "hyp", original.data = nhanes2
)
```

# 5 Plot against masked true values 
In general, we do not know the true values of the missing data, thus we can only plot the imputed values versus the observed data. However, if we happen to know the true values (e.g., in cases where it is generated by simulation), we may compare them to the imputed values.

Let us generate a simple dataset `full.df` and create 30% missing values in variable `norm1` and `norm2` under the MCAR mechanism. We then impute the `MCAR.df` dataset with `mixgb()`. 
```{r}
N <- 1000
norm1 <- rnorm(n = N, mean = 1, sd = 1)
norm2 <- rnorm(n = N, mean = 1, sd = 1)
y <- norm1 + norm2 + norm1 * norm2 + rnorm(n = N, mean = 0, sd = 1)
full.df <- data.frame(y = y, norm1 = norm1, norm2 = norm2)
MCAR.df <- createNA(data = full.df, var.names = c("norm1", "norm2"), p = c(0.3, 0.3))

mixgb.data <- mixgb(data = MCAR.df, m = 5, nrounds = 10)
```

If the true data is known, it can be specified in the argument `true.data` in the visual diagnostic functions. The output will then have an extra panel called `MaskedTrue`, which shows values originally observed but intentionally made missing.
```{r}
plot_hist(
  imputation.list = mixgb.data, var.name = "norm1",
  original.data = MCAR.df, true.data = full.df
)
plot_box(
  imputation.list = mixgb.data, var.name = "norm2",
  original.data = MCAR.df, true.data = full.df
)
plot_2num(
  imputation.list = mixgb.data, var.x = "norm1", var.y = "y",
  original.data = MCAR.df, true.data = full.df
)
```
